---
title: An Ode to the Unsung Functions of Purrr
author: Christina Fillmore
date: '2021-02-28'
slug: an-ode-to-the-unsung-functions-of-purrr
categories:
  - tidy-data
tags:
  - purrr
summary: |
  Exploring the wonderful uses for keep and discard, the "vector filtering functions" 
---
```{r, echo=FALSE}
library(purrr)
library(stringr)
```

When most people think of the purrr package they think of map, or maybe flatten if they are really into lists. But, I am here to make the case, that keep and discard are the most under valued functions in purrr. 

The origin of my love for keep and discard started about two years ago, when I was still somewhat new to the tidyverse. I was writing an internal package and I needed to "filter a vector". It seemed strange to me there wasn't an obvious solution for this problem; filter is one of the founding functions of dplry, how could it not exist!? At this point, while I knew how to do it in base R, I felt the uncompromising need of the newly converted, to write exclusively in the tidyverse. So, after asking a few friends, I got to googling and still couldn't find anything. And, as is the way, I stumbled upon keep and discard while looking through the documentation for a totally different question. Which, has since lead me to my current believe thatnkeep and discard are the most unsung functions in purrr. 

Keep and discard dramatically increase the readability of code, without sacrificing speed. All you need to do is to use a function that returns a vector of booleans, i.e. `is.na`. This can be a function you have created or a function from some other package. Here I created function to test if a value is even or not. 
```{r}
is_even <- function(x){
  x %% 2 == 0 
}

```
The `keep` function keeps all TRUE elements and `discard` removes all TRUE elements from the vector. 
```{r}
vec <- c(1:10)
#Returns only even numbers
vec %>% keep(is_even)
#Returns only odd numbers
vec %>% discard(is_even)
```
To do the same thing in base R you would need to call `vec` twice, which can make it difficult to read. 
```{r}
# Base method
vec[is_even(vec)]
```

The other great thing about about `keep` and `discard` is they work just as well with lists as vectors. These functions paired with an anonymous function is super helpful when sorting through complex lists.
```{r}
dummy_list <- list(a = mtcars, b = iris, c = NULL)

head(dummy_list)
```

In this dummy example list, I have 2 different datasets and a null input. If I wanted to remove the null value I can just use discard like I would on a vector. 
```{r}
dummy_list %>% discard(is.null) %>% head()
```
But, if I need something more complicated like, I need all datasets to have at least 100 observations I could use keep with an anonymous function. 
```{r}
dummy_list %>% 
  keep(function(x){
    # Removes null and small datasets 
    !is.null(x) && nrow(x) > 100
  }) %>% 
  head()
```

I hope this little post ensures `keep` and `discard` get a little bit more attention. Cause I find they really do vastly increase the reability of code.  